---
title: "Class 9: Candy Mini Project"
author: "Erin McTavish PID: A17300519"
format: pdf
toc: TRUE
---

## Background

In this mini-project, you will explore FiveThirtyEight’s Halloween Candy dataset.

We will use lots of **ggplot** some basic stats, correlation analysis and PCA to make sense of the landscape of US candy - something hopefully more relatable than the proteomics and transcriptomics work that we will use these methods on throughout the rest of the course

## Data Import

Our dataset is a CSV file so we use `read.csv()` peak with head()

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```
> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```


> Q3. What is your favorite candy (other than Twix) in the dataset and what is it's winpercent value?

```{r}
candy["Milky Way", ] $winpercent
```

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Milky Way") |> 
  select(winpercent)
```

> Q4. What is the winpercent value for "Kit Kat"?



```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Kit Kat") |> 
  select(winpercent)
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Tootsie Roll Snack Bars") |> 
  select(winpercent)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes! Winepercent.

```{r}
library("skimr")
skim(candy)
```

> Q7. What do you think a zero and one represent for the `candy$chocolate` column?

Zero: candy does not have chocolate. 1: candy has chocolate. 

## Exploratory analysis

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```


```{r}
library(ggplot2)
 ggplot(candy, aes(x=winpercent)) + 
   geom_histogram(biwidth = 3)
```
> Q9. Is the distribution of winpercent values symmetrical?

No

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

1. Find all chocolate candy
2. Get their winpercent values
3. Find the mean
4. Find all fruit candy
5. Get their winpercent values
6. Find the mean
7. Compare the two
```{r}
choc.candy <- candy[candy$chocolate == 1,]
choc.win <- choc.candy$winpercent
mean(choc.win)
```
```{r}
fruit.win <- candy[candy$fruity == 1,]$winpercent
mean(fruit.win)
```


> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```
It is statistically significant because the p-value is < 0.5.

## Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?

```{r}
y <- c("y", "a", "z")
sort(y)
```

```{r}
y
order(y)
```

```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,], 5)
```



>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[ord.ind,], 5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values

```{r}
ggplot(candy) +
  aes(winpercent, row.names(candy)) + 
  geom_col() +
  ylab("")
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent? 

```{r}
ggplot(candy) +
  aes(winpercent,
      reorder (row.names(candy), winpercent)) + 
  geom_col() +
  ylab("")
```


```{r}
ggplot(candy) +
  aes(winpercent,
      reorder (row.names(candy), winpercent, fill=chocolate)) + 
  geom_col() +
  ylab("")
```

We need a custom color vector

```{r}
my_cols <- rep("black", nrow(candy))
my_cols [candy$chocolate==1] <- "chocolate"
my_cols[candy$bar==1] <- "brown"
my_cols [candy$fruity==1] <- "yellow"
my_cols
```
```{r}
ggplot(candy,
  aes(winpercent, reorder (rownames(candy), winpercent))) + 
  geom_col(col=my_cols) +
  ylab("")
```

>Q17. What is the worst ranked chocolate candy?

Sixlets

>Q18. What is the best ranked fruity candy?

Starburst

##Taking a look at pricepercent

```{r}
library(ggrepel)

my_cols[candy$fruity==1] <- "red"

# How about a plot of win vs price

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
 geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese’s miniatures have high winpercent but low pricepercent

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip, Nestle Smarties, Ring pop, Hershey’s Krackel, Hershey’s Milk Chocolate are the 5 most expensive and least popular

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

  
  
## Exploring the correlation structure
  
```{r}
cij <- cor(candy)
```
```{r}
library(corrplot)
corrplot(cij)
```
>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity.

>Q23. Similarly, what two variables are most positively correlated?
  
  Sugarpercent and winpercent.
  

## Principal Component Analysis (PCA)


```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
Score plot... 

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x= PC1, y= PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 3)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

       
```{r}
ggplot(pca$rotation) +
  aes(PC1,
      reorder(row.names(pca$rotation), PC1)) +
  geom_col()
```

```{r}
#library(plotly)
#ggploty(p)
```

>Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

Fruity, pluribus and hard candies are picked up in the strongly positive direction. This was also highlighted in the correlation plot.  

>Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

Winning candies are mostly chocolate based because they have higher association with winpercent. 


  