---
title: "Class 18 Investigating Pertussis Resurgence"
author: "Erin McTavish PID: A17300519"
format: html
---


# Background
Pertussis (whopping cough) is a common lung infection caused by the bacteria B. Pertussis. It can infect anyone but is most deadly for infants (under 1 year old).


# CDC tracking data
The CDC tracks the number of Pertussis cases: https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

```{r, echo=FALSE}
cdc <- data.frame(
year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L,2024L,2025L),
cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063,22538,21996)
       )
```

>Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line()
```


>Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept =1946, col="blue", lty=2) +
  geom_vline(xintercept = 1996, col="red", lty=2) +
  geom_vline(xintercept=2020, col= "gray", lty=2)
```
Cases are decreasing from 1946 to 1996. After the wP vaccine was introduced, causes decreased dramatically, plateauing in the 70s-80s. 

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

When the switch to the aP vaccine occurs, things initially did not change much, but we are now seeing sparatic outbreaks in the early 2000s-present potentually due to immunity decreasing/pathogens adapting to the vaccine/less people are being vaccinated/the aP vaccine could be less effective. Moreover, there may be a waning efficacy of the aP vaccicne, as time passes, the vaccine becomes less effective in your body.

## Exploring CMI-PB data

The CMI-PB project < https://www.cmi-pb.org/ > mission of CMI-PB is to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination.

Basically, make available a large dataset on the immune respomnse to Pertussis. They use a "booster" vaccination as a proxy for Pertussis infection. 

They make their data available as JSON format API, we can read thie into R with the `read_json()` function from the **jsonlite** package: 


```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", 
                     simplifyVector =TRUE)
head(subject)
```
>Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```


>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```


>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```
>Q. Is this representative of the US population?

Absolutely not. By race, there is a lack of Alaska Natives, African American/Black, and Native Hawaiian individuals. 

We can now read mor etables from the CMI-PB database

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector =TRUE)
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector =TRUE)
```
```{r}
head(specimen)
```

```{r}
head(ab_titer)
```


>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(lubridate)
library(dplyr)

dob_col <- names(subject)[grepl("birth|dob", names(subject), ignore.case = TRUE)][1]

subject$age <- today() - ymd(as.character(subject[[dob_col]]))

ap <- subject %>% filter(infancy_vac == "aP")
round(summary(time_length(ap$age, "years")))

wp <- subject %>% filter(infancy_vac == "wP")
round(summary(time_length(wp$age, "years")))
```


>Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


To make sense of all this data we need to "join" (a.k.a. "merge" or "link") all these tables together. Only then will you know that a given Ab measurement (from the `ab_titer` table) was collected on a certain date (from the `speciment` table) from a certain wP or aP individual subject (from the `subjecy table). 

>Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

```{r}
# Or use wilcox.test() 
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```

We can use **dplyr** and the `*join()` family of functions to do this. 

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```
Let's do one more `inner_join()` to join the `ab_titer` with all our `meta` data. 

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```



>Q9 pt2. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
# Complete the API URLs...
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE) 

# Join specimen and subject tables
meta <- left_join(specimen, subject)

dim(meta)
head(meta)
```


>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)

dim(abdata)
```


>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```


>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$antigen)
```

Let's focus on IgG isotype 

```{r}
igg <- abdata |>
  filter(isotype=="IgG")

head(igg)
```

>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
    aes(MFI_normalised, antigen) +
    geom_boxplot()
```


>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others? 

The antigens "PT", "FIM2/3" and "FHA" appear to have the widest range of values. These are all components of the aP vaccine. These antigens have higher levels of IgG antitbody recognizing over time because you are being vaccinated by the aP booster shot and your immune system is recognizing and responding to it!  

>Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()

filter(igg, antigen=="PT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```
```{r}
filter(igg, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```


>Q16. What do you notice about these two antigens time courses and the PT data in particular?

The control antigen remains constant and low throughout all visits. The PT antigen increases drastically after the booster vaccines indicating a prevalent immune response. 

>Q17. Do you see any clear difference in aP vs. wP responses?

```{r}
ggplot(igg) +
    aes(MFI_normalised, antigen, col=infancy_vac) +
    geom_boxplot()
```
The wP group has higher PT antibodies than the aP group. Moreover, the wP group shows a stronger immune response to the vaccine booster than the aP group. 

```{r}
ggplot(igg) +
    aes(MFI_normalised, antigen) +
    geom_boxplot()+
  facet_wrap(~infancy_vac)
```


Is there a difference with time (ie.) before vs after booster shot?

```{r}
ggplot(igg) +
    aes(MFI_normalised, antigen) +
    geom_boxplot()
    facet_wrap(~visit)
```



>Q18. Does this trend look similar for the 2020 dataset?

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```



```{r}
library(bio3d)
ali <- read.fasta("PART2Findagene.fst")
```

```{r}
str(ali)
```

```{r}
ali_mat <- ali$ali
dim(ali_mat)
head(ali_mat)
```


```{r}
library(bio3d)
library(pheatmap)

ali <- read.fasta("PART2Findagene.fst")
ali_mat <- ali$ali

# pairwise identity matrix (0 to 1)
id_mat <- seqidentity(ali_mat, normalize = TRUE)

# optional: convert identity to distance
dist_mat <- 1 - id_mat

pheatmap(
  dist_mat,
  main = "Pairwise prolactin sequence distance heatmap"
)
```

```{r}
a1 <- pdb.annotate("1RW5")
a2 <- pdb.annotate("3EW3")
a3 <- pdb.annotate("2Q98")
```

```{r}
a1[, c("structureId", "experimentalTechnique", "resolution", "source")]
```

```{r}
a2[, c("structureId", "experimentalTechnique", "resolution", "source")]
```

```{r}
a3[, c("structureId", "experimentalTechnique", "resolution", "source")]
```
```{r}
names(hits$hit.tbl)
```

```{r}
hits$hit.tbl[, c("subjectids","identity","evalue")]
```

















